---
title: "README"
output: 
  rmarkdown::github_document:
    pandoc_args: --webtex
header-includes:
   - \usepackage{amsmath, amssymb, amsthm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayesian Additive Regression Trees for Multivariate skewed responses

This repository contains code to implement the methodology described in the paper "Bayesian Additive Regression Trees for Multivariate Skewed Responses", by Um et al (2021+).

Note that this package uses the primary functions from [`SoftBART`](https://github.com/theodds/SoftBART) to incorporate the SoftBART model as a component.

## Installation

This repository includes two packages: `skewBART` and `MultiskewBART`. The packages can be installed with the `devtools` package:

```{r, eval = FALSE}
library(devtools) 
devtools::install_github(repo='Seungha-Um/skewBART', subdir='/skewBART') 
devtools::install_github(repo='Seungha-Um/skewBART', subdir='/MultiskewBART', force = TRUE) 
```
## Documentation

Vignette is available at


## Status of developement

The code provided here is being actively developed and only being provided for research purposes. 

## Usage

#### skewBART model (univariate response)

The `skewBART` packages provides a nonparametric regression approach for univariate skewed responses using Bayesian additive regression trees (BART).

The following is an example on “Friedman’s example”:

```{r, eval = FALSE}
## Load library
library(skewBART)

## Generate fake data
sim_fried <- function(N, P, alpha, sigma) {
  lambda <- alpha * sigma/sqrt(1+alpha^2)
  tau <- sigma/sqrt(1+alpha^2)
  X <- matrix(runif(N * P), nrow = N)
  mu <- 10 * sin(pi * X[,1] * X[,2]) + 20 * (X[,3] - 0.5)^2 + 10 * X[,4] + 5 * X[,5]
  Z <- abs(rnorm(N, mean=0, sd=1) )
  Y <- mu + lambda * Z + rnorm(N, mean=0, sd=sqrt(tau))
  EY <- mu + lambda * sqrt(2/pi)
  return(list(X = X, Y = Y, EY = EY, mu = mu, Z=Z, tau = tau, lambda = lambda))
}

## Simulate dataset
set.seed(12345)
c(X,Y,EY,mu,Z,tau,lambda) %<-% sim_fried(250, 100, 5, 1)
c(test_X,test_Y,test_EY,test_mu,test_Z,test_tau,test_lambda)  %<-% sim_fried(250, 100, 5 ,1)
```

The predictors are scaled to lie in [0,1] using quantile normalization and the response Y is scaled to mean 0 and standard deviation 1 :

```{r, eval = FALSE}
X_norm <- quantile_normalize(X, test_X)$X
test_X_norm <- quantile_normalize(X, test_X)$test_X
Y_scaled <- scale(Y)
c(Y_dims, center_Y, scale_Y) %<-% attributes(Y_scaled)
```

The function `Hypers` produces a list of the hyperparameters of the model. 

```{r, eval = FALSE}
hypers <- Hypers(X = X_norm, Y = Y_scaled)
```

We can then fit the model as 

```{r, eval = FALSE}
## Fit the model
fit <- skewBART(X = X_norm, Y = Y_scaled, test_X = test_X_norm, tree_num=200, iter= 5000, burn =2500)

## Compare the quality of the fit
rmse <- function(x,y) sqrt(mean((x-y)^2))

rmse(fit$y_hat_train, EY)
rmse(fit$y_hat_test, test_EY)
```


#### MultiskewBART model (multivariate response)

The `MultiskewBART` packages provides a nonparametric regression approach for multivariate skewed responses using Bayesian additive regression trees (BART).


```{r, eval = FALSE}
## Load library
library(MultiskewBART)

## Generate fake data
sim_data_multi <- function(N, P, lambda, tau, rho) {
  X <- matrix(runif(N * P), nrow = N)
  mu <- 10 * sin(pi * X[,1] * X[,2]) + 20 * (X[,3] - 0.5)^2 + 10 * X[,4] + 5 * X[,5] 
  Z <- cbind(lambda[1] * abs(rnorm(N)), lambda[2] * abs(rnorm(N)))
  Sigma <- matrix(c(tau[1], sqrt(tau[1]*tau[2])*rho, sqrt(tau[1]*tau[2])*rho, tau[2]), 2, 2)
  Err <- mvrnorm(n=N, mu=c(0,0), Sigma = Sigma)
  Y <- cbind(mu, mu) + Z + Err
  EY <- rbind(mu, mu) + lambda * sqrt(2/pi)
  return( list(X = X, Y = Y, EY=EY, mu = mu, lambda = lambda, tau=tau, Z= Z, Sigma = Sigma) )
}

## Simulate dataset
set.seed(12345)
c(X,Y,EY,mu,lambda,tau,Z,Sigma) %<-% sim_data_multi(250, 5, c(2,3), c(1,1), 0.5)
c(test_X,test_Y,test_EY,test_mu,test_lambda,test_tau,test_Z,test_Sigma) %<-% sim_data_multi(100, 5, c(2,3), c(1,1), 0.5)

## preprocess data
X_norm <- quantile_normalize(X, test_X)$X
test_X_norm <- quantile_normalize(X, test_X)$test_X
Y_scaled <- scale(Y)
center_Y <- attributes(Y_scaled)$`scaled:center`
scale_Y <- attributes(Y_scaled)$`scaled:scale`

## Create a list of the hyperparameters of the model. 
hypers <- Hypers(X = X_norm, Y = Y_scaled)
```

We can then fit the model as 

```{r, eval = FALSE}
## Fit the model
fit <- MultiskewBART(X = X_norm, Y = Y_scaled, test_X = test_X_norm, tree_num=200, iter= 2500, burn =5000) 

## Compare the quality of the fit
rmse <- function(x,y) sqrt(mean((x-y)^2))
rmse(fit$y_hat_train, t(EY))
rmse(fit$y_hat_test, t(test_EY))
```








